Is high social class always beneficial for survival? A study of northern Sweden 1801--2013 \footnote{Short title, "Is high social class beneficial for survival?"} \footnote{Funded by the Swedish Research Council under grants 349-2008-6592 and 421-2012-892.}

### 2018-10-02 popstud_open.Rmd
cosmetics

---
title: Is high social class always beneficial for survival? A study of northern Sweden 1801--2013 \footnote{Short; "Is high social class beneficial for survival?"} \footnote{Funding:This work was founded by the Swedish Research Council under grants No 349-2008-6592 No 421-2012-892}
author: Sören Edvinsson \footnote{email soren.edvinsson@umu.se, orcid 0000-0001-7439-002X (corresponding author)}
  and Göran Broström \footnote{email goran.brostrom@umu.se, orcid 0000-0002-2865-7894}
date: Centre for Demographic and Ageing Research, Umeå University, Umeå, Sweden
output:
  bookdown::pdf_document2:
    citation_package: natbib
    keep_tex: yes
    number_sections: yes
    toc: no
    toc_depth: 2
  bookdown::html_document2:
    number_sections: yes
    toc: yes
    toc_depth: 2
    toc_float: yes
  bookdown::word_document2:
    toc: yes
    toc_depth: 2
citation_package: natbib
classoption: titlepage
biblio-style: apalike
documentclass: article
fontsize: 12pt
header-includes:
- \usepackage[utf8]{inputenc}
- \usepackage{graphicx}
- \usepackage{endfloat}
- \usepackage[doublespacing]{setspace}
- \usepackage{sectsty}
- \subsectionfont{\normalfont\itshape}
bibliography: mybib
abstract: "Focusing on two regions in northern Sweden 1801--2013, we challenge common
  notions of the assumed advantage in survival of belonging to a high social class.
  The issue is analysed according to gender and age group (adults and elderly) and
  in relation to the development of economic inequality. The results show that high
  social class is not always favourable for survival. Men in the elite category, particularly
  in working age, had higher mortality compared to others during a large part of the
  studied period; a male mortality class reversal appears at a surprisingly late date,
  while the social gradient among women conforms to the expected pattern. We suggest
  that health-related behaviour is decisive not only in later but earlier phases of
  the mortality transition as well. The results implicate that the association between
  social class and health is more complex than is assumed in many of the dominant
    theories in demography and epidemiology.\\par\n\\textbf{Keywords:} Class reversal
  in mortality; Cox regression; Gender differences; Inverse probability weighting;
  Mortality transition; Northern Sweden; Social inequality; 19th to 21 century."

@@

Figure \@ref(fig:plottaalla) shows the distribution
(per cent) of exposure time according to class for men and women,
respectively (see also Appendix). The farmer category was dominant until the middle of the 20th centuries but have become marginal in the latest periods, particularly among women. The category of workers is fairly stable over time. The highest class has increased substantially, albeit from a very low level
in the 19th century. The increase in academics in connection to the
establishment of Umeå University explains much of this increase in later time. Furthermore,
we observe a rather high percentage of missing social class in the early 19th century.
For men the proportion is somewhat higher in the age group 65 and above, those who were retired,
but still low enough to be used in our analyses (results not shown). If we disregard the very first period,
however, there were very few with occupation missing until 1950.

@@

The proportion with missing values among women is a larger problem, however. Corresponding to what we found for men, the
proportion without social position is very low until the middle of the 20th
century, reflecting the way women's social position is defined. However, the frequency of missing social class is quite high during particularly the first two periods of data from the Linnaeus database---from 1976. Since female social class for this period is defined
according to own occupation, the consequence of many women being outside
the workforce is that the proportion of missing increases. For the later periods,
the results are fairly complete.

@@

The social classes are compared by the *expected years of life
before 100* for a person aged 65, and the *relative risk of dying before age 65* for a person aged 40, respectively. The reason for using different comparative measures for the two age groups is the fact that mortality is very low in the younger age group compared to the older, at least during recent periods.

@@

```{r cumswtabout, message = FALSE, cache = FALSE, results='asis'}
##knitr::kable(dr, caption = "Significance tests, ages 40-64.", digits = c(0, 1, 2, 4))
print(xtable(dr,label = "tab:cumswtab", caption = "ANOVA, ages 40--64."),
      type = "latex", comment = FALSE)
```

@@

```{r cumswtaboldout, cap = "Table", cache = FALSE, results='asis'}
##knitr::kable(drold, caption = "Single term deletions, ages 65-99.", digits = c(0, 1, 2, 4))
##drold
print(xtable(drold,label = "tab:cumswtabold", caption = "ANOVA, ages 65--99."),
      type = "latex", comment = FALSE)
```

@@

# References

@@

The exposure in years for men and women respectively are shown in Table \@ref(tab:womenexp), and Table \@ref(tab:menexp).

```{r exposuretab, message = FALSE, cache = FALSE, results = 'asis'}
##expo <- aggregate(vb[, c("exposure")], by = vb[, c("hisclass", "period", "sex")], FUN = sum)
ex <- with(vb, tapply(exposure, list(period, hisclass, sex), sum))
print(xtable(ex[,,2], label = "tab:womenexp", caption = "Exposure in years by social class and period, women.", digits = 0), 
      type = "latex", comment = FALSE)
```

```{r exposuretab2, message = FALSE, cache = FALSE, results = 'asis'}
print(xtable(ex[,,1], label = "tab:menexp", caption = "Exposure in years by social class and period, men.", digits = 0), 
      type = "latex", comment = FALSE)
```

>>>>

---
title: Is high social class always beneficial for survival? A study of northern Sweden 1801--2013 \footnote{Short title, "Is high social class beneficial for survival?"} \footnote{Funded by the Swedish Research Council under grants 349-2008-6592 and 421-2012-892.}
author: Sören Edvinsson \footnote{email soren.edvinsson@umu.se, orcid 0000-0001-7439-002X (corresponding author).}
  and Göran Broström \footnote{email goran.brostrom@umu.se, orcid 0000-0002-2865-7894.}
date: Centre for Demographic and Ageing Research, Umeå University, Umeå, Sweden
output:
  bookdown::pdf_document2:
    citation_package: natbib
    keep_tex: yes
    number_sections: yes
    toc: no
    toc_depth: 2
  bookdown::html_document2:
    number_sections: yes
    toc: yes
    toc_depth: 2
    toc_float: yes
  bookdown::word_document2:
    toc: yes
    toc_depth: 2
citation_package: natbib
classoption: titlepage
biblio-style: apalike
documentclass: article
fontsize: 12pt
header-includes:
- \usepackage[utf8]{inputenc}
- \usepackage{graphicx}
- \usepackage{endfloat}
- \usepackage[doublespacing]{setspace}
- \usepackage{sectsty}
- \subsectionfont{\normalfont\itshape}
bibliography: mybib
abstract: "Focusing on two regions in northern Sweden 1801--2013, we challenge common
  notions of the assumed advantage in survival of belonging to a high social class.
  The issue is analysed according to gender and age group (adults and elderly) and
  in relation to the development of economic inequality. The results show that high
  social class is not always favourable for survival. Men in the elite category, particularly
  in working age, had higher mortality compared to others during a large part of the
  studied period; a male mortality class reversal appears at a surprisingly late date,
  while the social gradient among women conforms to the expected pattern. We suggest
  that health-related behaviour is decisive not only in later but earlier phases of
  the mortality transition as well. The results implicate that the association between
  social class and health is more complex than is assumed in many of the dominant
    theories in demography and epidemiology.\\par\n\\textbf{Keywords:} Class reversal
  in mortality; Cox regression; Gender differences; Inverse probability weighting;
  Mortality transition; Northern Sweden; Social inequality; 19th to 21 century."

@@

Figure \@ref(fig:plottaalla) shows the distribution
(per cent) of exposure times according to class for women and men (see also Tables \@ref(tab:womenexp) and \@ref(tab:menexp), respectively, in the Appendix). The farmer category was dominant until the middle of the 20th centuries but have become marginal in the latest periods, particularly among women. The category of workers is fairly stable over time. The highest class has increased substantially, albeit from a very low level
in the 19th century. The increase in academics in connection to the
establishment of Umeå University explains much of this increase in later time. Furthermore,
we observe a rather high percentage of missing social class in the early 19th century.
For men the proportion is somewhat higher in the age group 65 and above, those who were retired,
but still low enough to be used in our analyses (results not shown). If we disregard the very first period,
however, there were very few with occupation missing until 1950.

@@

The proportion of missing values among women is a larger problem, however. Corresponding to what we found for men, the
proportion without social position is very low until the middle of the 20th
century, reflecting the way women's social position is defined. However, the frequency of missing social class is quite high during particularly the first two periods of data from the Linnaeus database---from 1976. Since female social class for this period is defined
according to own occupation, the consequence of many women being outside
the workforce is that the proportion of missing increases. For the later periods,
the results are fairly complete.

@@

The social classes are compared by the *expected years of life before 100* for a person aged 65, and the *relative risk of dying before age 65* for a person aged 40. The reason for using different comparative measures for the two age groups is the fact that mortality is very low in the younger age group compared to the older, at least during recent periods.

@@

```{r cumswtabout, message = FALSE, cache = FALSE, results='asis'}
##knitr::kable(dr, caption = "Significance tests, ages 40-64.", digits = c(0, 1, 2, 4))
print(xtable(dr,label = "tab:cumswtabout", caption = "ANOVA, ages 40--64."),
      type = "latex", comment = FALSE)
```

@@

```{r cumswtaboldout, cap = "Table", cache = FALSE, results='asis'}
##knitr::kable(drold, caption = "Single term deletions, ages 65-99.", digits = c(0, 1, 2, 4))
##drold
print(xtable(drold,label = "tab:cumswtaboldout", caption = "ANOVA, ages 65--99."),
      type = "latex", comment = FALSE)
```

@@

@@

The exposure times in years for women and men are shown in Tables \@ref(tab:womenexp) and \@ref(tab:menexp), respectively.

```{r exposuretab, message = FALSE, cache = FALSE, results = 'asis'}
##expo <- aggregate(vb[, c("exposure")], by = vb[, c("hisclass", "period", "sex")], FUN = sum)
ex <- with(vb, tapply(exposure, list(period, hisclass, sex), sum))
print(xtable(ex[,,2], label = "tab:womenexp", caption = "Exposure time in years by social class and period, women.", digits = 0), 
      type = "latex", comment = FALSE)
```

```{r exposuretab2, message = FALSE, cache = FALSE, results = 'asis'}
print(xtable(ex[,,1], label = "tab:menexp", caption = "Exposure time in years by social class and period, men.", digits = 0), 
      type = "latex", comment = FALSE)
```
